---
title: "01 - Initial Analysis"
author: "Ryszard Szymański"
date: "5/24/2020"
output: html_document
---

```{r setup, include=FALSE}
library(patchwork)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
options(knitr.table.format = "html") 
```

## Introduction
The goal of this notebook is to perform an initial analasis of the CRAN package network which will lead to a number of experiment ideas for future work. Two different types of package networks have been created:

* One based on the `Suggests` relation
* One based on both the `Depends` and `Imports` relation

## Analysis
A basic summary of graph statistics is presented in the table below:
```{r}
drake::readd(graphs_summary) %>% 
  knitr::kable() %>% 
  kableExtra::kable_styling(full_width = FALSE)
```

As we are investigating package dependencies only the in degrees will be considered. The histogram of in degrees for both graphs are depicted in the following pictures:
```{r, out.width="100%"}
drake::readd(degree_hist_plot_suggests_graph) +
  drake::readd(degree_hist_plot_depends_graph)
```

Both graphs follow a power distribution and are on the less dense side. The difference between those graphs is that one is a DAG and the other is not, however this is not a surprise as the Depends graph should be a DAG while the suggests graph might contain crossdependcies between popular document and testing frameworks.


